I find the technology behind great products fascinating, but I'm even more fascinated by the conditions that lead to that technology.\\
\par
I don't feel in any way responsible for the greatness of \doom{}.  Technologically, that was all Carmack, and anything that came close in my code was only due to his influence.\\
\par
In the course of struggling to keep up with Carmack, I would sometimes fall asleep in the office.  I remember hearing that at least for a window there, Carmack felt guilty that I was working so hard.  Knowing his intense work ethic, perhaps that put a little more spring in his already indefatigable step.\\
\par
What I do know is that I had joined an already well-oiled development team in the form of Carmack, Romero, Adrian, and Kevin.  This wasn't their first rodeo.\\
\par
I later learned the mother of their invention was a harsh time constraint at Soft Disk, a company that sold a subscription of curious demos on diskette.  They wanted to start making games instead of just toy applications, and their manager allowed them, but only if they could deliver a game every two months like clockwork.\\
\par
Back then, there were no game engines.  In fact, DOS was not a complete operating system, and you had to finish writing the drivers for your game to work.  So this was an incredibly harsh time constraint, and it forced them into doing constant triage on what they wanted to achieve with each game.\\
\par
I was so convinced that this painful constraint is what led to the team's impressive work on \doom{}, that years later, I would teach a university class inspired by id's path to success.  The other teachers warned me not to be nice to the students, or they would take advantage of me.  Relishing an acting challenge, these were my first words to the class: "My name is Dave Taylor.  I'm a working producer, and I don't have time for you.  You owe me a shipped game every week, and if you don't ship, you'll get an F."\\
\par
Three classes and 127 games later, almost all of my students went on to jobs in the game industry, no mean feat with a game design degree, and many of them credit the painful time constraints of the class.  Terror works.  I recommend it.\\
\par
Which brings me to our newest source of terror.  I am not popular amongst my peers for having a low opinion of games.  I consider what we make an opiate for the mind with about as much redeeming value.  I met one too many \doom{} fans who expressed their enthusiasm in how it threatened their GPA's, their relationships with their significant others, and their employment.  That used to be funny and flattering until I integrated the area under the curve.\\
\par
Scalar money commerce does not formally motivate us to make games that are good for you, just to make money, which unfortunately makes us complicit in the Holocene Extinction now threatening the planet.  We need to start changing behaviors quickly, which means we desperately need games to help us change those behaviors.\\
\par
Terrified?  Good.  Get to work!\\
\par
        =-ddt->\\
\thispagestyle{plain} % Trick to remove header
