The success of the PC version and its mind numbing sales figure made it an extremely desirable title for any console publisher. From 1994 to 1997, \doom{} was ported to the six major systems of the era with more or less success.\\
\par
% \begin{itemize}
%    \item 1994: Atari Jaguar and the Sega 32X.
%    \item 1995: Super Nintendo and Sony Playstation.
%    \item 1996: 3DO.
%    \item 1997: Sega Saturn.
% \end{itemize}
% INSER ARRAY HERE: Console, Year, Authors, Avg framerate, price, CPUS, RAM.
% \bigskip

% \begin{figure}[H]
% \centering  
% \begin{tabularx}{\textwidth}{ L{1}  R{1} R{0.5} R{0.5} R{0.5} R{0.5} }
%   \toprule
%    \textbf{Year} & \textbf{System} & \textbf{\#CPUs} & \textbf{RAM} & \textbf{Resolution} & \textbf{Price} \\
%   \toprule 
%             1994 & Atari Jaguar.   &  5              & 2 MiB.       &  320x240            & BLA\\
%             1994 & Sega 32X        &  3              & ISA-8        &  8                  & BLA\\
%             1995 & Super Nintendo  &                 & ISA-16       & 13                  & BLA\\
%             1995 & Sony Playstation&                 & VLB          & 24                  & BLA\\
%             1996 & 3DO             &                 & VLB          & 24                  & BLA\\
%             1997 & Sega Saturn     &                 & VLB          & 24                  & BLA\\
%    \toprule
%  \end{tabularx}
% \caption{Ports of \doom to video game console from 1995 to 1997}
% \end{figure}

Four years during which the "console war" was raging, with consoles ranked by generation via their "bitness". Third generation, 8-bit based, NES and Sega Master System had long disappeared. The fourth, 16-bit generation, made of Nintendo's SNES, TurboGrafx-16 and Sega's Genesis was reaching end of life. The fifth, 32-bit, generation 
was starting to appear featuring Sony's Playstation and Sega Saturn with marketers trying to play on words with systems branded 64-bit like the Nintendo 64 or the Jaguar\footnote{After that, consumers either got tired or the "bits" or realized the silliness of the whole nomenclature. Bitness was forgotten.}. In retrospect, it was a rich period, prone to hardware innovation which contrast vividly when compared to 2018 uniform world of Sony vs Microsoft where system only differ by their name.\\
\par
The architecture of the engine based on a core with "just" sub-system to implement may lead to conclude it was an easy task to get \doom{} to run on a console. This feeling could not be further from the truth. From design trade-offs due to restricted resources, to crazy schedules, all versions have a unique and rich story to tell.\\
\par
Technically, the common problem to solve was to deal with less memory than originally intended. PCs minimum requirement was to have 4 MiB installed on the machine. It was of course not possible to ask customers to add more RAM to their console. Developers sometimes had to deal with as little as 512 KiB of RAM which was eight times less than the original version.\\
\par
The second technical difficulty was to dealing with exotic hardware. The PC was designed around one "big iron" CPU while consoles were a made of a constellation of processors.












\section{Jaguar (1994)}
\subfile{port_jaguar} 




\section{Sega 32X (1994)}
\subfile{port_x32} 






\section{Super Nintendo (1995)}
\subfile{port_snes}








\section{Playstation 1 (1995)}
\subfile{port_psx}










\section{3DO (1996)}
\subfile{port_3do}




\section{Saturn (1997)}
\subfile{port_saturn}

\fixme{Watch https://www.youtube.com/watch?v=784MUbDoLjQ before closing this chapter.}
