This is the second \textit{Game Engine Black Book}. It picks up right where the first one ended with the release of Wolfenstein 3D in May 1992. It carries on all the way up to December 1993 with id Software's second breakthrough of the 90s, \doom{}.\\ %in the world of PC gaming
\par
 Like its predecessor, this volume attempts to describe in great detail both the hardware and the software of the era. It opens a window back in time peeking over the engineering used to solve the various problems id Software encountered during the eleven months it took them to ship their next title.\\% which resulted in what is universally considered one of the best game of all times.\\
\par
It may seem odd to write a book about a game twenty-five years after its release. After all, who would be interested in seemingly outdated technology found in extinct hardware running obsolete operating systems? Given the success of the first \textit{Black Book}, it turns out many. Whether readers are into history, nostalgia, engineering, or even philosophy, it seems there is an edge for everyone.\\ 

\par
\doom{} has had such a profound and sustained impact that it has become part of modern history. It is an unquestionable milestone that entertained millions and catalyzed vocations. Because the source code was made available, programmers have learned the game engine's architecture with it. Because it was easy to modify and the tools were available, countless aspiring game makers had their first experiences designing or drawing assets. To this day, because it is such an icon, it is often the go-to title for hackers wanting to demonstrate their skills\footnote{Upon cracking the BitFi wallet in Aug 2018, the hacker team demonstrated it by running \doom{} on the device.}. From the MacBook Touchbar, to ATMs, CT scanners, watches, and even fridges, pretty much any piece of electronics has run \doom{} \footnote{ "Will it run \doom{}?" has become a common joke in the computer/gaming worlds. There is even a website, "itrunsdoom.tumblr.com", to provide the answer.}.\\
\par

It was a financial and critical success that reshaped the PC gaming industry\footnote{And even killed the Amiga. Source: "Commodore: The Amiga Years" by Brian Bagnall.}. During 1994 it received several awards, including \textit{Game of the Year} by both \textit{PC Gamer} and \textit{Computer Gaming World}, \textit{Award for Technical Excellence} from \textit{PC Magazine}, and \textit{Best Action Adventure Game Award} from the \textit{Academy of Interactive Arts \& Sciences}. With more than two million copies sold and an estimated 20 million shareware installations, at its height the phenomenon generated close to \$100,000 per day. Before the term was overtaken by "First Person Shooter" people talked about "Doom clones."\\
\par

 \rawdrawing{Doom_clone_vs_first_person_shooter}
 \par

There is also tremendous sentimental value. \doom{} is one of those titles that made an everlasting impression upon first contact. Those who were able to play it upon release or shortly after are still able to remember the circumstance under which they first saw it running. It is an exhilarating feeling to learn the internals of something once deemed magical.\\
 \par
 \vspace{-12pt}
Beyond the nostalgia, and this is the most important reason this book is relevant, the making of \doom{} is the ever-repeating story of inventors, engineers, and builders gathered around a common vision. There was no clear path from where id Software was to where they wanted to be -- only the certainty that nobody else had gone there before. They worked days and nights, slept on the floors, and waded across rivers to make their dreams come true.\\
\par
 The making of \doom{} summarizes well how achieving a colossal task breaks down to a thousand small things done right. This is the story of a group of dreamers who combined skills, dedication and good fortune, resulting in a breathtaking combination of technology, artwork and design.\\
\par



% You may disagree with the values of these "old" things. Some people prefer to sail with the wind, rarely looking back. But even to them this book could turn out to be a useful engineering map someday.\\
% \par
 To narrate this wonderful adventure, the black book had to respond to two seemingly orthogonal constraints. On the one hand, to be able to stand alone without need for supplemental information or cross-references. On the other hand, to avoid boring faithful readers with content already visited in the series. The middle ground was to allow people who had read about Wolfenstein 3D to get more out of this book without making it a necessity.\\
 \par
 Topics which would have been interesting to re-visit, such as the architecture of the VGA hardware, DOS TSRs, 386 Real-Mode, PC Speaker sound synthesis, the PIC and PIT, DDA algorithms and a few others are mentioned but not extensively described since they were part of \textit{Game Engine Black Book: Wolfenstein 3D}. This trade-off allowed reaching the target, which was a book around 400 pages that can be held in one hand while sipping a cup of tea.\\
\par
A few liberties were taken with regard to code samples. Due to the restricted real estate of the paper version, code sometimes had to be slightly modified to fit. Other times, in order to introduce complexity progressively and not overwhelm the reader, portions of the code in functions were removed. Some code samples are from the original source code before it was cleaned up during the open sourcing effort so it may differ from what you can find on \cw{github.com}. Rest assured the semantics and spirit remain intact.\\
\par 
This book is the fruit of an exercise inspired by Nicolas Boileau who reportedly stated: "Whatever we well understand, we express clearly". It is also the volume I wish someone else had written so I could just have purchased it (I am quite lazy).\\
\par
I hope you will enjoy reading it!\\
\par
-- Fabien Sanglard\\
Occasional Link to the Past\\
\par
Sunnyvale, CA\\
\monthyeardate\today\\
