\section{Music}
Like in their previous games, id asked Bobby Prince Jr. to create the music and some of the many audio effects. He was contacted before the game was up and running. To get a grasp of the ambiance he was only given the "Doom" Bible, a document authored by Tom Hall which acted as a design doc describing the tone of the game.\\
\par

\fq{Much of what was in Doom Bible never appeared in the game, but it set a mood for starting on the project. Within a few months of receiving that document, I had roughed out a lot of music and most of what turned out to be final sound effects.}{Bobby Prince, Retro Gamer 44.}
\par
Even though id Software set a general direction, it did not prevent Bobby from innovating and taking initiative.\\
\par
\fq{The id Software development team originally wanted me to do nothing but metal songs for DOOM. I did not think that this type of music would be appropriate throughout the game, but I roughed out several original songs and also created MIDI sequences of some cover material. This was before any level design and was before most of the artwork had been created. As the game came together, the guys at id saw that this type of music was not appropriate for many of the levels in DOOM. Thinking that this would be the case, I had also roughed out a lot of ambient moody background music, much of which ended up in the game. This song ("At Doom's Gate") was one of the first of its type that I wrote. I heard it as being on a level that went by real fast. As it turns out, John Romero (who placed all of the songs on the levels) decided it was a perfect song for the first level.}{Bobby}
\par
Popular metal bands of the early nineties such as Metallica, Believer, Slayer, Alice in Chains, AC/DC, and especially Pantera served as sources of inspiration. 




\section{Sounds}
For the audio effects, the "The General Series 6000" pack was purchased from Sound Ideas. The royalty-free set of 5 CDs was of such high quality that its samples are still used to this day. An attuned ear will recognize the Doom's door opening sound in the music video of "Body Movin'" by the Beastie Boys\footnote{5:09 mark.} and in the Doctor Who TV series.\\
\par
\trivia{If you ever wondered what kind of mixture of animal growling sounds could have possibly been the origin of the Imp dying sample, it was a plain and simple camel.}\\
\par


